---
title: "DATA 609 - HW 7"
author: "Betsy Rosalen"
date: "December 5, 2021"
output:
    html_document:
        theme: cerulean
        code_folding: show
        df_print: paged
        toc: true
        toc_depth: 2
        toc_float:
            collapsed: false
            smooth_scroll: true
        css: ./style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
options(scipen=999, digits = 4)
library(tidyverse)
library(kableExtra)
library(caret)
library(e1071)
set.seed(42)
```

## Ex. 1

### Use the `svm()` algorithm of the `e1071` package to carry out the support vector machine for the `PlantGrowth` data set.  Then discuss the number of support vectors/samples.  [Install the `e1071` package in R if needed.]


```{r ex1}
p <- PlantGrowth
cbind(p[1:10,],p[11:20,],p[21:30,])
```

```{r ex1.2}
p_svm <- svm(group ~ weight, data = p)
summary(p_svm)
```

```{r ex1.3}
p_pred <- predict(p_svm, newdata = p)
table(p_pred)
```

```{r ex1.4}
confusionMatrix(p_pred, p$group)
```

  
## Ex. 2

### Do a similar SVM analysis as that in the previous question using the `iris` data set.  Discuss the number of support vectrs/samples.  


```{r ex2}
i <- iris
i
```

```{r ex2.2}
i_svm <- svm(Species ~ ., data = i)
summary(i_svm)
```

```{r ex2.3}
i_pred <- predict(i_svm, newdata = i)
table(i_pred)
```

```{r ex2.4}
confusionMatrix(i_pred, i$Species)
```


## Ex. 3

### Use the `iris` data set (or any other data set) to select 80% of the samples for training `svm()`, then use the rest 20% for validation.  Discuss your results.  

```{r ex3}
set.seed(42)
index <- createDataPartition(iris$Species, p = 0.80, list = FALSE)
train <- iris[index, ]
test <- iris[-index, ]
```

```{r ex3.2}
svm_iris <- svm(Species ~ ., data = train)
summary(svm_iris)
```

```{r ex3.3}
pred_iris <- predict(svm_iris, newdata = test)
table(pred_iris)
```

```{r ex3.4}
confusionMatrix(pred_iris, test$Species)
```



